Type-Safe Client-Server Communication in Web Applications

### 2014-02-27 paper.tex
Finish up chatbox revisited section.

\begin{itemize}
  \item The programming model is synchronous, giving the programmer a simple,
        linear view of the program flow, eliminating the need to program with
        callbacks and continuations.
  \item Side-effecting code is explicitly designated to run on either the
        client or the server using the type system while pure code can be
        shared by both. Additionally, both client and server code may lift
        general IO computations, allowing for safe IO code reuse within the
        confines of the client or server designated code sections.
  \item Client-server network communication is handled through statically typed
        RPC function calls, extending the reach of Haskell's type checker over
        the network and giving the programmer advance warning when she uses
        network services incorrectly or forgets to update communication code
        as the application's internal protocol changes.
  \item Our model takes the view that the client side is the main driver when
        developing web applications and accordingly delegates the server to
        the role of a computational and/or storage resource, tasked with
        servicing client requests rather than driving the program.
  \item The implementation is built as a library on top of the GHC and Haste
        Haskell compilers, requiring little to no compiler support. Programs
        are compiled twice; once with Haste and once with GHC, to produce the
        final client and server side code respectively.
\end{itemize}

@@

Therein lies the secret to restricting \lstinline!liftServerIO! to the
server: the only function capable of turning a \lstinline!Useless! value into
a useful one lives in the server monad. Any client side code is thus free to
completely ignore executing computations lifted using \lstinline!liftServerIO!;
since the result of such a computation is never observable on the client,
the client has no obligation to even produce such a value.
Figure \ref{lst:good-state} shows how to make proper use of server-side state.

@@

\begin{code}
srvHello :: Useless RcptList -> Server ()
srvHello uRcpts = do
  recipients <- mkUseful uRcpts
  sid <- getSessionID
  rcptMVar <- liftIO CC.newEmptyMVar
  liftIO $ CC.modifyMVar recipients $ \cs -> do
    return ((sid, rcptMVar):cs ,())
\end{code}

@@

\begin{code}
srvSend :: Useless RcptList -> String -> Server ()
srvSend uRcpts message = do
    rcpts <- mkUseful uRcpts
    liftIO $ do
      recipients <- CC.withMVar rcpts return
      mapM_ (CC.forkIO . deliver message) recipients
  where
    deliver :: String -> Recipient -> IO ()
    deliver message (_, rcptMVar) =
      CC.putMVar rcptMVar message
\end{code}

@@

\begin{code}
srvAwait :: Useless RcptList -> Server String
srvAwait uRcpts = do
  rcpts <- mkUseful uRcpts
  sid <- getSessionID
  liftIO $ do
    recipients <- CC.withMVar rcpts return
    case find ((== sid) . fst) recipients of
      Just (_, mv) -> CC.takeMVar mv
      _            -> fail "Unregistered session; aborting"
\end{code}

@@

@@

Our model is based on a simple blocking concurrency model rather than explicit
continuations, is well suited for use with a GUI programming style based on
communicating processes with local state, and requires essentially no
modification of existing tools or compilers.

>>>>

\begin{itemize}
  \item The programming model is synchronous, giving the programmer a simple,
        linear view of the program flow, eliminating the need to program with
        callbacks and continuations.
  \item Side-effecting code is explicitly designated to run on either the
        client or the server using the type system while pure code can be
        shared by both. Additionally, both client and server code may lift
        general IO computations, allowing for safe IO code reuse within the
        confines of the client or server designated code sections.
  \item Client-server network communication is handled through statically typed
        RPC function calls, extending the reach of Haskell's type checker over
        the network and giving the programmer advance warning when she uses
        network services incorrectly or forgets to update communication code
        as the application's internal protocol changes.
  \item Our model takes the view that the client side is the main driver when
        developing web applications and accordingly delegates the server to
        the role of a computational and/or storage resource, tasked with
        servicing client requests rather than driving the program. While it is
        entirely possible to implement a server-to-client communication channel
        on top of our model, we believe that choosing one side of the
        heterogenous client-server relation as the master helps keeping the
        program flow linear and predictable.
  \item The implementation is built as a library on top of the GHC and Haste
        Haskell compilers, requiring little to no compiler support. Programs
        are compiled twice; once with Haste and once with GHC, to produce the
        final client and server side code respectively.
\end{itemize}

@@

Therein lies the secret to restricting \lstinline!liftServerIO! to the
server: the only function capable of turning a \lstinline!Useless! value into
a useful one lives in the server monad. Any client side code is thus free to
completely ignore executing computations lifted using \lstinline!liftServerIO!;
since the result of a server lifted computation is never observable on the
client, the client has no obligation to even produce such a value.
Figure \ref{lst:good-state} shows how to make proper use of server-side state.

@@

\begin{code}
srvHello :: Useless RcptList -> Server ()
srvHello uRcpts = do
  recipients <- mkUseful uRcpts
  sid <- getSessionID
  liftIO $ do
    rcptMVar <- CC.newEmptyMVar
    CC.modifyMVar recipients $ \cs ->
      return ((sid, rcptMVar):cs ,())
\end{code}

@@

\begin{code}
srvSend :: Useless RcptList -> String -> Server ()
srvSend uRcpts message = do
    rcpts <- mkUseful uRcpts
    liftIO $ do
      recipients <- CC.withMVar rcpts return
      mapM_ (CC.forkIO . deliver message) recipients
  where
    deliver msg (_, rcptMVar) = CC.putMVar rcptMVar msg
\end{code}

@@

\begin{code}
srvAwait :: Useless RcptList -> Server String
srvAwait uRcpts = do
  rcpts <- mkUseful uRcpts
  sid <- getSessionID
  liftIO $ do
    recipients <- CC.withMVar rcpts return
    case find ((== sid) . fst) recipients of
      Just (_, mv) -> CC.takeMVar mv
      _            -> fail "Unregistered session!"
\end{code}

@@

If the caller's session identifier could not be found in the
recipient list, it has for some reason not completed its handshake with the
server. If this is the case, we simply drop the session by throwing an error;
the session will be automatically cleaned up on the server side and an exception
is thrown to the client.

Having implemented our three server operations, all that's left is to tie them
to the client. We start by creating an \lstinline!MVar! to hold the server
state and exporting our three server operations.

\begin{code}
appMain :: App Done
appMain = do
  recipients <- liftServerIO $ CC.newMVar []

  hello <- export $ srvHello recipients
  awaitMsg <- export $ srvAwait recipients
  sendMsg <- export $ srvSend recipients

  runClient $ do
    withElems ["log","message"] $ \[log,msgbox] -> do
      onServer hello
\end{code}

Notice that the \lstinline!recipients! list is passed to our three server
operations \emph{before} they are exported; since \lstinline!recipients! is
created on the server and inaccessible to the client side, it is not possible
to pass it over the network as an RPC argument. Even if it were possible,
passing server-private state back and forth over the network would be highly
undesirable.

The \lstinline!withElems! function is part of Haste's DOM manipulation library;
it locates references to the DOM nodes with the given identifiers, in this case
the variable \lstinline!log! is bound to the node with the identifier ``log'',
and \lstinline!inp! is bound to the node identifier by ``msg''. These are the
same DOM nodes that were referenced in our original example, and refer
to the chat log window and the text input field respectively. After locating
all the needed UI elements, the client proceeds to register itself with the
server's recipient list using the \lstinline!hello! export.

\begin{code}
      let recvLoop chatlines = do
            setProp log "value" $ unlines chatlines
            message <- onServer awaitMsg
            recvLoop (message : chatlines)
      fork $ recvLoop []
\end{code}

The \lstinline!recvLoop! function perpetually asks the server for new messages
and updates the chat log whenever one arrives. Note that unlike the
\lstinline!onmessage! callback of the Javascript version of this example,
\lstinline!recvLoop! is acting as a completely self-contained process with
linear program flow, keeping track of its own state and only reaching out to
the outside world to write its state to the chat log whenever necessary. As
the \lstinline!awaitMsg! function blocks until a message arrives,
\lstinline!recvLoop! will make exactly one iteration per received message.

\begin{code}
      msgbox `onEvent` OnKeyPress $ \13 -> do
        msg <- getProp msgbox "value"
        setProp msgbox "value" ""
        onServer (sendMsg <.> msg)
\end{code}

This is the final part of our program; we set up an event handler to clear the
input box and send its contents off to the server whenever the user hits return
(character code 13) while the input box has focus. Finally, all that is left
is to actually launch our application.

Using this approach, creating web applications with strongly typed
communication becomes easy. The discerning reader may be slightly annoyed at
the need to extract the contents from \lstinline!Useless! values at each point
of use. Indeed, in a simple example such as this, the source clutter caused by
this becomes a disproportionate irritant. Fortunately, most web applications
tend to have more complex client-server interactions, reducing this overhead
significantly.

@@

Our model is based on a simple blocking concurrency model rather than explicit
continuations which is well suited for use with a GUI programming style based on
communicating processes with local state, and requires essentially no
modification of existing tools or compilers.
